Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-06-04 PAPER.tex
Mon Jun  4 11:48:25 EDT 2018

\subsection{Signal Loss}
\label{sec:SiglossOverview}

Signal loss can arise in a variety of ways in the analysis pipeline, such as by fitting a polynomial during spectral calibration, applying a delay-domain filter, or deriving weights from data and applying them to itself. Here we focus on signal loss associated with 
% we're not considering a general problem, but are really just focused on the empirical QE and its discontents, so let's say that
the use of an empirically estimated covariance matrix with the ``optimal quadratic estimator'' formalism.
This loss was significantly 
%applying a weighting matrix to data, a loss that can be significant depending on the choice of weighting and one that was previously 
underestimated in the \citetalias{ali_et_al2015} analysis.

% Put all of the (O)QE / ICW theory here, rather than sprinkled throughout this section.

@@

The goal of the power spectrum analysis is to produce an unbiased estimator of the EoR power spectrum in the presence of both noise and foreground emission.  Prior to power spectrum estimation, the data will often have been prepared to have as little foreground as possible by some method of subtraction, so this foreground emission may appear either directly (because it was not subtracted) or as a residual to some subtraction process not in the power spectrum domain.  If we possess a good estimate of the total covariance of the data, including both the desired signal nd any contaminants, then the ``optimal quadratic estimator'' formalism provides a method of producing a minimum variance, unbiased estimator of the

%Driven by the need to mitigate foreground bias, PAPER's previous analyses use a weighting method that aims to down-weight foregrounds. This weighting is applied to data, which is then propagated into a final estimator using the power spectrum estimation technique of quadratic estimators (QE) as done in 
\citet{liu_tegmark2011}, \citet{dillon_et_al2013a}, \citet{liu_et_al2014a}, \citet{liu_et_al2014b}, \citet{trott_et_al2012}, \citet{dillon_et_al2014}, \citet{dillon_et_al2015}, \citet{switzer_et_al2015}, and \citet{trott_et_al2016}. Before showing 
how signal loss can arise when using different weighting matrices, we first summarize QE as performed in the PAPER 
analysis.

@@

@@

We illustrate the way in which the quadratic estimator, Equations \eqref{eq:qhat} and \eqref{eq:phat}, is intended to estimate the power spectrum of EoR in the presence of contamination by other terms by considering two simpler, but similar, cases of estimating the variance of data.  We have specifically chosen a toy model where the data covariance is diagonal, as indeed we expect the EoR signal to be.  We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and an undesired contaminant $\upsilon_i$
\begin{equation}
\Delta_i = \sigma_i + \upsilon_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \bm{\sigma \sigma}^T \rangle = s \mathbf{I}_{N \times N} \equiv \mathbf{S}
\end{equation}
and we wish to estimate $s$.  The contaminant in this case has a similar structure for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\langle \upsilon_i \rangle = 0; \; 
\langle \upsilon^2_i \rangle = u; \; 
\langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{I}_{N \times N} \equiv \mathbf{U}; {\rm and}~ 
\langle \sigma_i \upsilon_j \rangle = 0
\end{equation}
With the covariance matrix given by $\C = \mathbf{S} + \mathbf{U}$, the estimator for $s$ using the optimal quadratic estimator prescription is
\begin{equation}
\label{eq:ICWFormula}
\hat{s} = \half {\F}^{-1} \left( \bm{\Delta}^T \invC \PDeriv{\C}{s}  \invC \bm{\Delta} \right) 
\end{equation}
where $\F$ is the Fisher matrix.  
The expectation of the estimator is then
\begin{equation}
\langle \hat{s} \rangle = \frac{\langle \bm{\Delta^T \Delta} \rangle}{N} = s + u
\end{equation}
In this case, the quadratic estimator returns an entirely reasonable result, but, since the covariance of the undesired signal has the same form as the desired signal, that is,
$\PDeriv{\mathbf{S}}{s} = \PDeriv{\mathbf{U}}{u} =  \PDeriv{\C}{s}$, we find that when {\it the covariance structure of the contaminant is identical to the signal}, the information available to the quadratic estimator to distinguish the two is degenerate, and  is only possible to achieve an unbiased estimate of $s$ by using some additional information to estimate the additive bias due to $u$ and subtract it.  In fact, if one uses the standard formula \citep[e.g.][, Equation 6]{dillon14} for the additive bias
\begin{equation}
b = \half \F^{-1} \Tr\left[ (\C - \mathbf{S}) \invC \PDeriv{\C}{s} \invC \right]
\end{equation}
one indeed finds $b = u$.

@@

We can again calculate the estimator given by \eqref{eq:ICWFormula} explicitly by using the Sherman-Morrison identity to invert the covariance matrix.  Defining
\begin{equation}
\xi \equiv \frac{u/s}{1+u/s}
\end{equation} 
we have
\begin{equation}
\invC  =   \frac{1}{s} \left(\I - \xi \mathbf{m m}^T \right)
\end{equation}
Proceeding to calculate the full estimator
\begin{equation}
\hat{s} = \half {\F}^{-1} \left(\bm \Delta^T \invC \PDeriv{\C}{s}  \invC \bm \Delta \right) = \frac{ \bm \Delta^T  (\I +  (\xi^2 - 2 \xi)  \mathbf{m m}^T)  \bm \Delta}{N + \xi^2 - 2 \xi} 
\end{equation}
The expectation is
\begin{equation}
\langle \hat{s} \rangle = 
s + \frac{1 - 2 \xi + \xi^2}{N + - 2 \xi + \xi^2} u 
\end{equation}
It is worth observing immediately that there is no multiplicative bias on $s$, and that the additive bias is strictly $< u/N$.

@@

We find the following things:
\begin{itemize}

>>>>

\subsection{Signal Loss}
\label{sec:SiglossOverview}
% JEA - We never totally nail down what ``signal loss'' is ...

Signal loss can arise in a variety of ways in the analysis pipeline, such as by fitting a polynomial during spectral calibration, applying a delay-domain filter, or deriving weights from data and applying them to itself. Here we focus on signal loss associated with 
% JEA - we're not considering a general problem, but are really just focused on the empirical QE and its discontents, so let's say that
the use of an empirically estimated covariance matrix with the ``optimal quadratic estimator'' formalism.
This loss was significantly 
%applying a weighting matrix to data, a loss that can be significant depending on the choice of weighting and one that was previously 
underestimated in the \citetalias{ali_et_al2015} analysis.

% JEA - Put all of the (O)QE / ICW theory here, rather than sprinkled throughout this section.

@@

%AAAICW

The goal of the power spectrum analysis is to produce an unbiased estimator of the EoR power spectrum in the presence of both noise and foreground emission.  Prior to power spectrum estimation, the data will often have been prepared to have as little foreground as possible by some method of subtraction, so this foreground emission may appear either directly (because it was not subtracted) or as a residual to some subtraction process not in the power spectrum domain.  If we possess a good estimate of the total covariance of the data, including both the desired signal and any contaminants, then the ``optimal quadratic estimator'' formalism provides a method of producing a minimum variance, unbiased estimator of the desired signal
% JEA - I cut this b/c I wanted to step back and explain a bit more clearly what we were actually hoping for 
%Driven by the need to mitigate foreground bias, PAPER's previous analyses use a weighting method that aims to down-weight foregrounds. This weighting is applied to data, which is then propagated into a final estimator using the power spectrum estimation technique of quadratic estimators (QE) as done in 
\citet{liu_tegmark2011}, \citet{dillon_et_al2013a}, \citet{liu_et_al2014a}, \citet{liu_et_al2014b}, \citet{trott_et_al2012}, \citet{dillon_et_al2014}, \citet{dillon_et_al2015}, \citet{switzer_et_al2015}, and \citet{trott_et_al2016}. Before showing 
how signal loss can arise when using different weighting matrices, we first summarize QE as performed in the PAPER 
analysis.

The optimal quadratic estimator prescription is
\begin{equation}
\label{eq:OQE}
\hat{p}_\alpha = \half ({\F}^{\alpha \beta})^{-1} \left( \bm{\Delta}^T \invC \PDeriv{\C}{p_\beta}  \invC \bm{\Delta} + \Tr\left[ (\C - \mathbf{S}) \invC \PDeriv{\C}{s} \invC \right] \right)
\end{equation}
where $\F$ is the Fisher matrix.

@@

In summary, we expect:
\begin{itemize}

In the following sections, we find

@@

We illustrate the way in which the quadratic estimator, Equations \eqref{eq:qhat} and \eqref{eq:phat}, is intended to estimate the power spectrum of EoR in the presence of contamination by other terms by considering two simpler, but similar, cases of estimating the variance of data.  We have specifically chosen a toy model where the data covariance is diagonal, as indeed we expect the EoR signal to be.  We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and an undesired contaminant $\upsilon_i$
\begin{equation}
\Delta_i = \sigma_i + \upsilon_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \bm{\sigma \sigma}^T \rangle = s \mathbf{I}_{N \times N} \equiv \mathbf{S}
\end{equation}
and we wish to estimate $s$.  The contaminant in this case has a similar structure for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\langle \upsilon_i \rangle = 0; \; 
\langle \upsilon^2_i \rangle = u; \; 
\langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{I}_{N \times N} \equiv \mathbf{U}; {\rm and}~ 
\langle \sigma_i \upsilon_j \rangle = 0
\end{equation}
With the covariance matrix given by $\C = \mathbf{S} + \mathbf{U}$, the estimator for $s$ using the quadratic part of Equation \ref{eq:OQE} is
\begin{equation}
\hat{s} = \frac{ \bm{\Delta^T \Delta}}{N} 
\end{equation}
and its expectation is
\begin{equation}
\langle \hat{s} \rangle = s + u
\end{equation}
In this case, the quadratic estimator returns an entirely reasonable result, but, since the covariance of the undesired signal has the same form as the desired signal, that is,
$\PDeriv{\mathbf{S}}{s} = \PDeriv{\mathbf{U}}{u} =  \PDeriv{\C}{s}$, we find that when {\it the covariance structure of the contaminant is identical to the signal}, the information available to the quadratic estimator to distinguish the two is degenerate, and  is only possible to achieve an unbiased estimate of $s$ by using some additional information to estimate the additive bias due to $u$ and subtract it.  In fact, computing the linear bias from Equation, \ref{eq:OQE}one indeed finds $b = u$.

@@

We can again calculate the estimator given by \eqref{eq:ICWFormula} explicitly by using the Sherman-Morrison identity to invert the covariance matrix.  Defining
\begin{equation}
\xi \equiv \frac{u/s}{1+u/s}
\end{equation} 
we have
\begin{equation}
\invC  =   \frac{1}{s} \left(\I - \xi \mathbf{m m}^T \right)
\end{equation}
Again calculating only the quadratic portion of the estimator,
\begin{equation}
\hat{s} = 
%\half {\F}^{-1} \left(\bm \Delta^T \invC \PDeriv{\C}{s}  \invC \bm \Delta \right) = 
\frac{ \bm \Delta^T  (\I +  (\xi^2 - 2 \xi)  \mathbf{m m}^T)  \bm \Delta}{N + \xi^2 - 2 \xi} 
\end{equation}
with expectation
\begin{equation}
\langle \hat{s} \rangle = 
s + \frac{1 - 2 \xi + \xi^2}{N + - 2 \xi + \xi^2} u 
\end{equation}
It is worth observing immediately that there is no multiplicative bias on $s$, and that the additive bias is strictly $< u/N$.

@@
